---
title: "Forecasting of a Groundwater Level Timeseries in Steenkoppies using Recurrent Neural Network (NARX)"
author: "Kirsty Gibson, Maximilian Nölscher"
date: "`r format(Sys.time(), '%d/%m/%Y')`"
output:
  html_document:
    theme: lumen
    highlight: tango
    toc: TRUE
    toc_float: TRUE
    collapse: FALSE
    toc_depth: 3
    collapsed: FALSE
    smooth_scroll: FALSE
    number_sections: TRUE
    self_contained: yes
    df_print: paged
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
```

```{r klippy, echo=FALSE, include=TRUE}
#klippy::klippy()
```

Source all external functions in folder functions
```{r}
purrr::map(
  list.files(
    path = './scripts/functions',
    pattern = "\\.R$",
    full.names = TRUE,
    recursive = TRUE
  ),
  source
)
```


Load Packages
```{r}
library(import)
import::from(readxl, read_excel)
import::from(janitor, clean_names, get_dupes)
import::from(forecast, nnetar, forecast, accuracy)
library(plotly)
# library(Metrics)
# library(doParallel)
library(timetk)
library(modeltime)
library(tidymodels)
library(tidyverse)
library(Metrics)
library(MLmetrics)
library(hydroGOF)
library(forecast)
library(dplyr)
library(recipes)
library(timetk)
library(devtools)

```

```{r}
devtools::install_github("business-science/modeltime")
```

# Data Import
```{r}
data <- read_excel("C:/Users/Kirsty/Desktop/Steenkoppies Data/1. GWL background decline/A2N0612/A2N0612.xlsx")
```

Show the imported dataframe
```{r}
data
```

# Minor Preparations

Clean column names
```{r}
data <- data %>% 
  clean_names()
```

Cast date column as date
```{r}
data <- data %>% 
  mutate(date = as.Date(date))
```



# Preprocessing
Convert dataframe into long format
```{r}
data <- data %>% 
  pivot_longer(cols = -date)
```

Plot the timeseries
```{r}
data %>% 
  ggplot(aes(x = date, 
             y = value)) + 
  geom_line(colour = "steelblue") +
  facet_wrap(~name, scales = "free_y", ncol = 1) +
  theme_minimal()
```

Convert back to wide format
```{r}
data <- data %>% 
  pivot_wider()
```


# Modelling

Now we can start modelling the prepared timeseries. We will do this using the tidymodels framework.
This helps us with data splitting (train, test, validation), hyperparameter tuning, normalization (scaling), resampling (timeseries cross-validation) and evaluating model performance. 

We start with data splitting.

## Data Splitting with `rsample`

The following figure shows the general resampling strategy
![Source: [Kuhn and Johnson (2019)](https://bookdown.org/max/FES/resampling.html)](figures/resampling.svg)

We start with the train-test-split by using 90% as training data and the residual 10% as testing data.

```{r}
train_test_splits <- data %>% 
  initial_time_split(prop = 0.9)
```

```{r}
data_train <- 
  train_test_splits %>% 
  training()
```

```{r}
data_test <- 
  train_test_splits %>% 
  testing()
```

Let's plot the train-test-split
```{r, fig.height=1.6, fig.width=10}
data_train %>% 
  mutate(split = "training") %>% 
  bind_rows(data_test) %>% 
  replace_na(list(split = "testing")) %>% 
  mutate(split = as.factor(split)) %>% 
  ggplot(aes(date, gwl)) +
  geom_line(aes(colour = split)) +
  theme_minimal() +
  scale_color_manual(values = c("training" = "#F8766D", "testing" = "#00BFC4")) +
  labs(colour = "Split") +
  scale_x_date(breaks = "2 years",
               labels = lubridate::year)
```


## Resampling with `rsample`

For defining the cross-validation split sizes we need to obtain some numbers on the sample size etc.
Get total number of observations aka timesteps of the training set
```{r}
n_samples_train <- 
  data_train %>% 
  nrow()
```

Define initial split size to 50% of the total number of observations
```{r}
n_initial <- 
  (n_samples_train * 0.5) %>% 
  floor()
```

Define the size of the following slices to a fifth of the remaining number of observations
```{r}
n_slices <- 5
```


```{r}
n_slice <- 
  ((n_samples_train - n_initial) / n_slices) %>% 
  floor()
```

Now we can define the resampling strategy
```{r}
resampling_strategy_cv5fold <- 
  data_train %>%
  time_series_cv(
    initial = n_initial,
    assess = n_slice,
    skip = n_slice,
    cumulative = TRUE
  )
```

Plot the resampling strategy
```{r, fig.width=10}
resampling_strategy_cv5fold %>% 
  tk_time_series_cv_plan() %>%  
  ggplot(aes(date, gwl)) +
  geom_line(aes(colour = .key)) +
  facet_wrap(~.id, ncol = 1) +
  theme_minimal() +
  labs(colour = "Split") +
  scale_x_date(breaks = "2 years",
               labels = lubridate::year)
```

## Preprocessing with `recipe`

All futher preprocessing steps such as normalization and feature engineering can be done using `recipe`
We only normalize the data in the following recipe
```{r}

steenkopies_recipe <-

  recipe(gwl ~ ., data = data_train) %>%

  step_normalize(all_numeric(), -date)
```

## Defining a Learner with `parsnip` and `modeltime`

Now we define a machine learning algorithm. We use a recurrent neural net (RNN) `nnetar()` from the `forecast` package which is implemented as `parsnip` model in the `modeltime` package.
This specific RNN is similar to the [NARX](https://www.mathworks.com/help/deeplearning/ug/design-time-series-narx-feedback-neural-networks.html;jsessionid=9c66d64c96648c267982f4c17e83) model in Matlab. 

 - We keep the hyperparameter seasonal_period fixed at 12 as 12 months is the annual seasonality
 - We keep the hyperparameter num_networks fixed at 20 as we expect not much impact of this hyperparameter
 - We tune all other hyperparameters by either using the default range or adapting it.
 

```{r}
tune_nnetar_model <-
  nnetar_reg(
    seasonal_period = 12,
    non_seasonal_ar = tune(),
    seasonal_ar = tune(),
    hidden_units = tune(),
    num_networks = 20,
    penalty = tune(),
    epochs = tune()
  ) %>%
  set_engine("nnetar", 
             scale.inputs = FALSE) %>%
  set_mode("regression")
```


## Parallelize Tuning

Parallel computation would be good, but I couldn't get it working after a first trial.
```{r}
# all_cores <- parallel::detectCores(logical = FALSE)
# 
# cluster <- makePSOCKcluster(all_cores)
# registerDoParallel(cluster)
```

```{r}
# stopCluster(cluster)
```


## Tuning

The setting of `n_levels` determines the number of values that are tried out for each hyperparameter. This causes an exponential growth of the number of models that need to be fitted.
`n_levels` = 3 for these five hyperparameters took approximately 30 minutes to calculate on an average PC. So this is a trade-off situation between number of tuned hyperparameter, computing time and number of hyperparameter combinations. Maybe one can try tuning a smaller subset of hyperparameters and let it run over the weekend with `n_levles` = 20.
```{r}
n_levels <- 1
```

Define tuning method. We use a regular grid.
```{r}
tune_grid <- grid_regular(
  non_seasonal_ar(range = c(1L, 5L)),
  seasonal_ar(range = c(1L, 5L)),
  hidden_units(),
  # num_networks(),
  penalty(),
  epochs(),
  levels = n_levels
)
```

Number of parameter combinations and consequently model runs:
```{r}
tune_grid %>% 
  nrow()
```


### Workflow
We put all this together in a workflow...
```{r}
nnetar_workflow <- 
  workflow() %>% 
  add_model(tune_nnetar_model) %>% 
  add_recipe(steenkopies_recipe)
```

### Model Fitting with Resamples
...and fit the models using the defined workflow. This is the core code chunk which can run for a long time.
```{r}
nnetar_resampling <- 
  nnetar_workflow %>% 
  tune_grid(
    resamples = resampling_strategy_cv5fold,
    grid = tune_grid)
    #metrics = metric_set(rmse, mae))
```

We choose a performance metrics (rmse or mae in this case)
```{r}
chosen_metric <- "rmse"
```

## Model Evaluation
We look at the influence of the tuned hyperparameters on the model performance.
```{r, fig.width=10, fig.height=6}
nnetar_resampling %>% 
  collect_metrics() %>% 
  pivot_longer(cols = c("non_seasonal_ar", "seasonal_ar", "hidden_units", "penalty", "epochs")) %>% 
  filter(.metric == chosen_metric) %>% 
  group_by(name, value) %>% 
  summarise(mean = mean(mean)) %>% 
  ggplot(aes(value, mean)) +
  geom_line(size = 1.5, 
            alpha = 0.6,
            colour = "#00BFC4") +
  geom_point(size = 2,
             colour = "#00BFC4") +
  facet_wrap(~ name, scales = "free_x", ncol = 1) +
  labs(y = "mean rmse") +
  theme_minimal()
```

We display the best model among all models
```{r}
nnetar_resampling %>%
  show_best(metric = chosen_metric)
```

We select the best model
```{r}
best_model <- 
  nnetar_resampling %>% 
  select_best(metric = chosen_metric)
```

We apply our workflow to the selected `best_model`
```{r}
final_workflow <- 
  nnetar_workflow %>% 
  finalize_workflow(best_model)
```

And we fit it to the whole training data split
```{r}
final_model <- 
  final_workflow %>% 
  fit(data = data_train)

```


We write the fitted model as modeltime_table
```{r}
models_tibble <- 
  final_model %>% 
  modeltime_table()
```

...and apply the model to the test data split
```{r}
calibration_tibble <- 
  models_tibble %>% 
  modeltime_calibrate(new_data = data_test)

```

We run the model on the test data split
```{r, fig.width=10}
best_model_forecast <- 
  calibration_tibble %>%
  modeltime_forecast(
    new_data = data_test,
    actual_data = data
  )
```

```{r}
extract_step_item <- function(recipe, step, item, enframe = TRUE) {
  d <- recipe$steps[[which(purrr::map_chr(recipe$steps, ~ class(.)[1]) == step)]][[item]]
  if (enframe) {
    tibble::enframe(d) %>% tidyr::spread(key = 1, value = 2)
  } else {
    d
  }
}

unnormalize <- function(x, rec, var) {
  var_sd <- extract_step_item(rec, "step_normalize", "sds") %>% dplyr::pull(var)
  var_mean <- extract_step_item(rec, "step_normalize", "means") %>% dplyr::pull(var)

  (x * var_sd) + var_mean
}

unnormalize(new_data,steenkopies_recipe, gwl)
```


We can now plot the models result
```{r}
best_model_forecast %>%
  plot_modeltime_forecast(.conf_interval_show = FALSE,
    .legend_max_width = 25 # For mobile screens
  )
```

We save the best model in `models_tibble`
```{r}
models_tibble %>% 
  write_rds("data_processed/models_tibble_best_model.Rds")
```

We plot the actual vs. predicted values of the test data split as scatter plot to get another impression of the model performance.
```{r}
actual_vs_prediction_data <- 
  best_model_forecast %>% 
  get_dupes(.index) %>% 
  select(.index, .key, .value) %>% 
  pivot_wider(names_from = ".key",
              values_from = ".value")

axis_limits <- 
  actual_vs_prediction_data %>% 
  pivot_longer(cols = -".index") %>% 
  pull(value) %>% 
  range()
  
actual_vs_prediction_data %>% 
  ggplot(aes(actual, prediction)) +
  geom_point() +
  coord_equal() +
  geom_abline(linetype = "dashed",
              colour = "grey",
              alpha = .6,
              size = 1) +
  xlim(axis_limits) +
  ylim(axis_limits) +
  theme_minimal()
```

